---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Andrew Friedman-Herring"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd()
library(tidyverse)
library(lubridate)
library(agricolae)
NTLphysics <- read.csv("../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")

NTLphysics$sampledate<- as.Date(NTLphysics$sampledate, format = "%m/%d/%y")
#2
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer: 
H0: Mean lake temperature in July does not change in depth across all lakes
Ha: 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
NTLphysics.july <- NTLphysics %>% 
  mutate(month = month(sampledate)) %>% 
  filter(month == 7) %>% 
  select(lakename, year4, daynum, depth, temperature_C) %>% 
  drop_na()

#5
ggplot(NTLphysics.july, aes(x = depth, y = temperature_C))+
  geom_point()+
  geom_smooth(method = "lm")+
  lims(y = c(0,35))+
  labs(x = "Depth (m)", y = "Temperature (*C)")

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: Overall, the tepmerature decreases with depth, then levels out at about 4 degrees C around 9 or
10 meters deep. The top couple of meters in the lake do not show this trend, likely due to wind and waves mixing 
the top layer of the water column. There is a decent spread in the data points, but the upper bounds on the data
points seems to show a relatively linear trend.



7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
NTLphysics.july.lm <- lm(NTLphysics.july$temperature_C ~ NTLphysics.july$depth)
summary(NTLphysics.july.lm)
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: Overall, 74% of the variability in temperature is explained by depth in this model, as per the R^2
statistic. This is found using 9501 degrees of freedom, and the p value is extremely low, meaning we can rule out
the null hypothosis that the temperature does not change with depth. According to the slope coefficient, for every
one meter of change in depth, the temperature will change 1.9 degrees.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
NTLphysics.july.lm.aic <- lm(NTLphysics.july$temperature_C ~ NTLphysics.july$year4 + 
                               NTLphysics.july$daynum + 
                               NTLphysics.july$depth)
step(NTLphysics.july.lm.aic)

#10
summary(NTLphysics.july.lm.aic)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of variables recommended by the model is all of the variables: Year, Depth, and Daynum.
This new model explains 70.4% of the variability in the data, which is a 1% improvement over the single variable model.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
NTLphysics.july.anova <- aov(data = NTLphysics.july, temperature_C ~ lakename)
summary(NTLphysics.july.anova)

NTLphysics.july.anova.lm <- lm(data = NTLphysics.july, temperature_C ~ lakename)
summary(NTLphysics.july.anova.lm)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, both models show that there is a statistically significant differnece in temperature in respect to
lake name. The p value for each test was on the order of 10^-16. 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
ggplot(NTLphysics.july, aes(x = depth, y = temperature_C, color = lakename))+
  geom_point(alpha = 0.5)+
  geom_smooth(method = "lm", se = FALSE)+
  lims(y = c(0,35))+
  labs(x = "Depth (m)", y = "Temperature (*C)")

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(NTLphysics.july.anova)
HSD.plot<- HSD.test(NTLphysics.july.anova, "lakename", group = TRUE)
HSD.plot
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Peter Lake falls into group C according to the Tukey HSD, which means its mean temperature is
statistically the same as Paul Lake and Ward Lake. No lake has a mean temperature that is statistically distinct
from all other lakes. Every group of means has at least two members in it. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could perform a two sample T-test to check if their means are equivalent. 


