---
output:
  pdf_document: default
  html_document: default
---
---==
title: "Assignment 09: Data Scraping"
author: "Andrew Friedman-Herring"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

# Total points:

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_09_Data_Scraping.Rmd") prior to submission.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1
getwd()
library(tidyverse)
library(rvest)
library(lubridate)

mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2019 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Change the date from 2020 to 2019 in the upper right corner.
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020>

Indicate this website as the as the URL to be scraped. (In other words, read the contents into an
`rvest` webpage object.)

```{r set.the.scraping.website}
#2
webpage <- read_html("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2019")

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Average Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to three separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values, with the first value being 36.0100.

```{r scrape.the.data}
#3
water.system.name <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
pswid <- webpage %>% 
  html_node("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
ownership <- webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
max.withdrawals.mgd <- webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in order. You can overcome this by creating a month column in the same order the data are scraped: Jan, May, Sept, Feb, etc...

5. Plot the max daily withdrawals across the months for 2020

```{r create.a.dataframe.from.scraped.data}
#4
Durham.WS <- data.frame("Month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
                        "Year" = rep(2019, 12),
                        "Max_Withdrawals_mgd" = as.numeric(max.withdrawals.mgd))

Durham.WS <- Durham.WS %>% 
  mutate("Date" = my(paste(Month, "-", Year)),
                    "Ownership" = !!ownership,
                    "PSWID" = !!pswid,
                    "Water System Name" = !!water.system.name)


#5
ggplot(Durham.WS, aes(x = Date, y = Max_Withdrawals_mgd))+
  geom_line()


```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site scraped**.

```{r construct.a.scraping.function}
#6.

scrape.it <- function(the_year,the_facility){
  the_base_url <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid='
  the_website <- read_html(paste0(the_base_url, the_facility, '&year=', the_year))
  
  water.system.name.tag <- "div+ table tr:nth-child(1) td:nth-child(2)"
  pswid.tag <- "td tr:nth-child(1) td:nth-child(5)"
  ownership.tag <- "div+ table tr:nth-child(2) td:nth-child(4)"
  max.withdrawals.mgd.tag <- "th~ td+ td"
  
  the_water.system.name <- the_website %>% html_nodes(water.system.name.tag) %>% html_text()
  the_pswid <- the_website %>% html_nodes(pswid.tag) %>% html_text()
  the_ownership <- the_website %>% html_node(ownership.tag) %>% html_text()
  the_max.withdrawals <- the_website %>% html_nodes(max.withdrawals.mgd.tag) %>% html_text()
  
  df_maxwithdrawals <- data.frame("Month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
                        "Year" = rep(the_year,12),
                        "Max_Withdrawals_mgd" = as.numeric(the_max.withdrawals)) %>% 
    mutate(Date = my(paste(Month, "-", Year)),
           PSWID = !!the_pswid,
           Water.system.name = !!the_water.system.name,
           Ownership = !!the_ownership)
  Sys.sleep(1)
  return(df_maxwithdrawals)
}


```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
df_durham_2015 <- scrape.it(2015, '03-32-010')

ggplot(df_durham_2015, aes(x = Date, y = Max_Withdrawals_mgd))+
  geom_line()

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares the Asheville to Durham's water withdrawals.

```{r fetch.and.plot.Ashville.2015.data}
#8
df_ashville_2015 <- scrape.it(2015,"01-11-010")
ggplot(df_durham_2015, aes(x = Date, y = Max_Withdrawals_mgd))+
  geom_line(data = df_durham_2015, aes(color = "Durham"))+
  geom_line(data = df_ashville_2015, aes(color = "Ashville"))

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

```{r}
#9 
year_range <- c(2010:2019)

Ashville_dfs <- lapply(X = year_range, 
                       FUN = scrape.it, 
                       the_facility= "01-11-010")

Ashville_2010_2019 <- bind_rows(Ashville_dfs)

ggplot(Ashville_2010_2019, aes(x = Date, y = Max_Withdrawals_mgd))+
  geom_line()+
  geom_smooth()
```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?

>Answer: Yes, the water usage was relatively stable from 2010-2016, and then has been increasing rapidly since 2016. 

